---
title: "Figure: latitudinal stability gradient"
output: html_document
date: "2025-12-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.width = 10)
```

```{r}
library(tidyverse)
library(FAVA)
library(patchwork)
library(broom)
```

```{r}
triton <- data.table::fread("../Data_raw/Data_Triton_v2_Maike_compressed/Triton_Binned_Trimmed.csv")

ecogroup <- data.table::fread("../Data_raw/Data_Triton_v2_Maike_compressed/Ecogroup.csv")

data_pal = PNWColors::pnw_palette("Bay", 3) %>% `names<-`(c("ecogroup", "morphogroup", "species"))

```

# Make exactly the same color palette for the species/ecogroups/morphgroups 

(we wont use these data sets otherwise)
```{r}
species_abundances_0.5my_alldeg = triton %>% 
  filter(`Macro.micro` == "Macroperforate") %>%
  transmute(species,
            age = round(age*2)/2, 
            abundance)  %>%
  group_by(age, species) %>%
  summarise(abundance = sum(abundance)) %>%
  pivot_wider(names_from = species, 
              values_from = abundance, 
              values_fill = 0)

ecogroup_abundances_0.5my_alldeg = triton %>% 
  left_join(ecogroup) %>%
  filter(`Macro.micro` == "Macroperforate") %>%
  transmute(ecogroup = `Ecogroup(s)`,
            age = round(age*2)/2, 
            abundance)  %>%
  group_by(age, ecogroup) %>%
  summarise(abundance = sum(abundance)) %>%
  filter(ecogroup !="") %>%
  pivot_wider(names_from = ecogroup, 
              values_from = abundance, 
              values_fill = 0)

morphogroup_abundances_0.5my_alldeg = triton %>% 
  left_join(ecogroup) %>%
  filter(`Macro.micro` == "Macroperforate") %>%
  transmute(morphogroup = factor(Morphogroup, levels = as.character(1:19), ordered=TRUE),
            age = round(age*2)/2, 
            abundance)  %>%
  group_by(age, morphogroup) %>%
  summarise(abundance = sum(abundance)) %>%
  filter(morphogroup !="") %>%
  pivot_wider(names_from = morphogroup, 
              values_from = abundance, 
              values_fill = 0)

# Normalize rows to generate relative abundances
species_relab_0.5my_alldeg = species_abundances_0.5my_alldeg
species_relab_0.5my_alldeg[,-1] = species_abundances_0.5my_alldeg[,-1] /rowSums(species_abundances_0.5my_alldeg[,-1])

ecogroup_relab_0.5my_alldeg = ecogroup_abundances_0.5my_alldeg
ecogroup_relab_0.5my_alldeg[,-1] = ecogroup_abundances_0.5my_alldeg[,-1] /rowSums(ecogroup_abundances_0.5my_alldeg[,-1])

morphogroup_relab_0.5my_alldeg = morphogroup_abundances_0.5my_alldeg
morphogroup_relab_0.5my_alldeg[,-1] = morphogroup_abundances_0.5my_alldeg[,-1] /rowSums(morphogroup_abundances_0.5my_alldeg[,-1])


comp_time_data = rbind(species_relab_0.5my_alldeg %>% arrange(age) %>% 
                         pivot_longer(cols = -1, names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Species"), 
                       
                       ecogroup_relab_0.5my_alldeg %>% arrange(age) %>% 
                         pivot_longer(cols = -1, names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Ecogroup"),
                       
                       morphogroup_relab_0.5my_alldeg %>% arrange(age) %>% 
                         pivot_longer(cols = -1, names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Morphogroup") ) %>% 
  filter(abundance > 0) %>% ungroup

comp_time_data$category = paste0(comp_time_data$data_type, comp_time_data$category)

comp_time_data$category =  factor(comp_time_data$category, ordered = TRUE, 
                                  c(paste0("Species", colnames(species_relab_0.5my_alldeg)[-1]),
                                    paste0("Ecogroup", colnames(ecogroup_relab_0.5my_alldeg)[-1]),
                                    paste0("Morphogroup", colnames(morphogroup_relab_0.5my_alldeg)[-1] %>% rev)) %>%
                                    rev)
# levels = comp_time_data %>% group_by(category)%>% 
#   summarize(total = sum(abundance)) %>% 
#   arrange(total) %>% select(category) %>% unlist)


pal_all = c(viridisLite::viridis(n=ncol(species_relab_0.5my_alldeg) -1) %>%
              `names<-`(filter(comp_time_data, data_type == "Species") %>% 
                          pull(category) %>% 
                          sort %>%
                          unique),
            viridisLite::viridis(n=ncol(ecogroup_relab_0.5my_alldeg) -1) %>%
              `names<-`(filter(comp_time_data, data_type == "Ecogroup")  %>% 
                          pull(category) %>% 
                          sort %>%
                          unique),
            viridisLite::viridis(n=ncol(morphogroup_relab_0.5my_alldeg) -1) %>%
              `names<-`(filter(comp_time_data, data_type == "Morphogroup") %>% 
                          pull(category) %>% 
                          sort %>%
                          unique))

```


# Analyze the data in bins 5 million years wide and 5 degress latitude wide

We break down the data into both spatial and temporal bins, whose definitions are visualized in the following plots. 

```{r 5 degree/million year bins}
# Define bins for the age and paleo latitude 
age_bins = data.frame(age = 0:66, 
                      age_bin = rep(seq(0,66,5),each=5)[1:67])
ggplot(age_bins, aes(x=age,y=age_bin)) +
  geom_line(aes(x=0:66, y=0:66), color="red")+
  geom_line() + geom_point() 

lat_bins = data.frame(pal.lat = -75:88, 
                      lat_bin = rep(seq(-76,88,5),each=5)[-165] + 1) %>% 
  mutate(lat_bin_label = paste0("[",lat_bin, ", ", lat_bin+4, "]"))
ggplot(data=lat_bins, aes(x=pal.lat,y=lat_bin)) +
  geom_line() + geom_point()+
  geom_line(data = data.frame(pal.lat=-76:88, lat_bin=-76:88), color="red", alpha=0.5)


# ------------------------------------------------------------------------------
# Make a table of species counts for each age_bin/lat_bin combo
species_abundances_5my_5deg = triton %>% 
  filter(`Macro.micro` == "Macroperforate") %>%
  transmute(species, 
            age = round(age), 
            pal.lat = round(pal.lat), 
            abundance)  %>%
  left_join(age_bins) %>%
  left_join(lat_bins) %>% 
  group_by(age_bin, lat_bin, lat_bin_label, species) %>%
  summarise(abundance = sum(abundance)) %>%
  pivot_wider(names_from = species, 
              values_from = abundance, 
              values_fill = 0)

# Remove the 15 age/lat combos with counts < 5
# sum(rowSums(species_abundances_5my_5deg[,-c(1:3)])<5)
species_abundances_5my_5deg = species_abundances_5my_5deg[rowSums(species_abundances_5my_5deg[,-c(1:3)])>=5,]

# Normalize rows to generate relative abundances
species_relab_5my_5deg = species_abundances_5my_5deg
species_relab_5my_5deg[,-c(1:3)] = species_abundances_5my_5deg[,-c(1:3)] /rowSums(species_abundances_5my_5deg[,-c(1:3)])


# ------------------------------------------------------------------------------
# Make a table of ecogroup counts for each age_bin/lat_bin combo
ecogroup_abundances_5my_5deg = triton %>% 
  filter(`Macro.micro` == "Macroperforate") %>%
  left_join(ecogroup) %>%
  transmute(ecogroup = `Ecogroup(s)`, 
            age = round(age), 
            pal.lat = round(pal.lat), 
            abundance)  %>%
  left_join(age_bins) %>%
  left_join(lat_bins) %>% 
  group_by(age_bin, lat_bin, lat_bin_label, ecogroup) %>%
  summarise(abundance = sum(abundance)) %>%
  filter(ecogroup !="") %>%
  pivot_wider(names_from = ecogroup, 
              values_from = abundance, 
              values_fill = 0)

# Remove the 15 age/lat combos with counts < 5
# sum(rowSums(ecogroup_abundances_5my_5deg[,-c(1:3)])<5)
ecogroup_abundances_5my_5deg = ecogroup_abundances_5my_5deg[rowSums(ecogroup_abundances_5my_5deg[,-c(1:3)])>=5,]

# Normalize rows to generate relative abundances
ecogroup_relab_5my_5deg = ecogroup_abundances_5my_5deg
ecogroup_relab_5my_5deg[,-c(1:3)] = ecogroup_abundances_5my_5deg[,-c(1:3)] /rowSums(ecogroup_abundances_5my_5deg[,-c(1:3)])

# ------------------------------------------------------------------------------
# Make a table of morphogroup counts for each age_bin/lat_bin combo
morphogroup_abundances_5my_5deg = triton %>% 
  filter(`Macro.micro` == "Macroperforate") %>%
  left_join(ecogroup) %>%
  transmute(morphogroup = Morphogroup, 
            age = round(age), 
            pal.lat = round(pal.lat), 
            abundance)  %>%
  left_join(age_bins) %>%
  left_join(lat_bins) %>% 
  group_by(age_bin, lat_bin, lat_bin_label, morphogroup) %>%
  summarise(abundance = sum(abundance)) %>%
  filter(morphogroup !="") %>%
  pivot_wider(names_from = morphogroup, 
              values_from = abundance, 
              values_fill = 0)

# Remove the 15 age/lat combos with counts < 5
# sum(rowSums(morphogroup_abundances_5my_5deg[,-c(1:3)])<5)
morphogroup_abundances_5my_5deg = morphogroup_abundances_5my_5deg[rowSums(morphogroup_abundances_5my_5deg[,-c(1:3)])>=5,]

# Normalize rows to generate relative abundances
morphogroup_relab_5my_5deg = morphogroup_abundances_5my_5deg
morphogroup_relab_5my_5deg[,-c(1:3)] = morphogroup_abundances_5my_5deg[,-c(1:3)] /rowSums(morphogroup_abundances_5my_5deg[,-c(1:3)])

# ------------------------------------------------------------------------------
# FILTER DATA SETS SO THAT EACH LAT BIN CONTAINS AT LEAST 10 YEAR BINS 
species_relab_5my_5deg_filtered = species_relab_5my_5deg %>%
                      filter(lat_bin %in% 
                               sort(unique(species_relab_5my_5deg$lat_bin))[
                                 table(species_relab_5my_5deg$lat_bin) >= 10]) 
ecogroup_relab_5my_5deg_filtered = ecogroup_relab_5my_5deg %>%
                      filter(lat_bin %in% 
                               sort(unique(species_relab_5my_5deg$lat_bin))[
                                 table(species_relab_5my_5deg$lat_bin) >= 10]) 
morphogroup_relab_5my_5deg_filtered = morphogroup_relab_5my_5deg %>%
                      filter(lat_bin %in% 
                               sort(unique(species_relab_5my_5deg$lat_bin))[
                                 table(species_relab_5my_5deg$lat_bin) >= 10]) 
```

```{r}
# Make a dataset that puts them all together
comp_time_lat = rbind(species_relab_5my_5deg_filtered %>% 
                         pivot_longer(cols = -c(age_bin, lat_bin, lat_bin_label), names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Species"), 
                       
                       ecogroup_relab_5my_5deg_filtered %>% 
                         pivot_longer(cols = -c(age_bin, lat_bin, lat_bin_label), names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Ecogroup"),
                       
                       morphogroup_relab_5my_5deg_filtered %>% 
                         pivot_longer(cols = -c(age_bin, lat_bin, lat_bin_label), names_to = "category", values_to = "abundance") %>% 
                         mutate(data_type = "Morphogroup") ) %>% 
  filter(abundance > 0) %>% ungroup

comp_time_lat$category = paste0(comp_time_lat$data_type, comp_time_lat$category)
comp_time_lat$category = factor(comp_time_lat$category, ordered=TRUE,
                                levels = pal_all %>% names)
comp_time_lat$lat_bin_label = factor(comp_time_lat$lat_bin_label, ordered=TRUE, 
                                     levels = lat_bins$lat_bin_label %>% unique)

# comp_time_lat$category =  factor(comp_time_lat$category, ordered = TRUE, 
#                                   c(paste0("Species", colnames(species_relab_5my_5deg_filtered)[-c(1,2)]),
#                                     paste0("Ecogroup", colnames(ecogroup_relab_5my_5deg_filtered)[-c(1,2)]),
#                                     paste0("Morphogroup", colnames(morphogroup_relab_5my_5deg_filtered)[-c(1,2)] %>% rev)) %>%
#                                     rev)
```

### Heterogeneity over time within each paleo latitude bin
```{r}
comp_time_lat_plot = ggplot(comp_time_lat,
                   aes(x = age_bin, y = abundance, fill = category, color = category)) + 
  geom_bar(stat = "identity") + 
  theme_bw() + 
  ggh4x::facet_grid2(data_type~lat_bin_label, 
                     strip = ggh4x::strip_themed(text_y=lapply(data_pal, function(col) element_text(color = col, face = "bold")))) + 
  theme(legend.position = "none", strip.background = element_blank()) + 
  scale_fill_manual(values = pal_all)+ 
  scale_color_manual(values = pal_all) + ylab("Relative abundance") + xlab("Million years ago") + 
  scale_x_continuous(limits = c(66.5,-0.5), breaks = c(60,30,0), trans = "reverse") + 
  scale_y_continuous(breaks = c(0,.5,1))

comp_time_lat_plot

# ggsave("../Figures/relab_time_latbin.png", comp_time_lat_plot, width = 41, height = 11, scale=0.3)
```

## Compute FAVA in sliding windows across time windows for each latitude bin

This analysis computes FAVA in sliding window four samples wide.

```{r}
species_age.windows = 
  FAVA::window_fava(relab_matrix = species_relab_5my_5deg_filtered,
                    window_size = 4,
                    K = 335,
                    group = "lat_bin_label",
                    time = "age_bin")

ecogroup_age.windows = 
  FAVA::window_fava(relab_matrix = ecogroup_relab_5my_5deg_filtered,
                    window_size = 4,
                    K = 18,
                    group = "lat_bin_label",
                    time = "age_bin")

morphogroup_age.windows = 
  FAVA::window_fava(relab_matrix = morphogroup_relab_5my_5deg_filtered,
                    window_size = 4,
                    K = 19,
                    group = "lat_bin_label",
                    time = "age_bin")


age.windows_all = morphogroup_age.windows$window_data %>%
  transmute(group, window_index, w1, w2, w3, w4, morphogroup = FAVA) %>% 
  left_join(ecogroup_age.windows$window_data %>% 
              transmute(group, window_index, w1, w2, w3, w4, ecogroup = FAVA)) %>% 
  left_join(species_age.windows$window_data %>% 
              transmute(group, window_index, w1, w2, w3, w4, species = FAVA)) %>%
  pivot_longer(cols = c(morphogroup, ecogroup, species), 
               names_to = "data", values_to = "FAVA")  %>% 
  pivot_longer(cols = c(w1, w2, w3, w4), names_to = "position", values_to = "mya")

age.windows_all$group = factor(age.windows_all$group, ordered=TRUE, levels = unique(lat_bins$lat_bin_label))


lat.stab.grad.panels = ggplot(age.windows_all %>% 
                                mutate(data = stringr::str_to_title(data)))+ 
  geom_smooth(method = "lm", aes(x = mya, y = FAVA), linewidth = 0.3, color = "black")  + 
  geom_line(aes(x = mya, y = FAVA, 
                color = data, 
                group = paste0(window_index, group, data, FAVA)),
            linewidth = 1, alpha=0.6) + theme_bw() + 
  ggh4x::facet_grid2(data~group, 
                     strip = ggh4x::strip_themed(text_y=lapply(data_pal, function(col) element_text(color = col, face = "bold"))))+ 
  theme(legend.position = "none", strip.background = element_blank()) + 
  scale_x_reverse(breaks = c(60, 30, 0))+ 
  scale_color_manual(values = data_pal %>% `names<-`(c("Ecogroup", "Morphogroup", "Species")), guide="none") + 
  xlab("Million years ago")
lat.stab.grad.panels

# ggsave("../Figures/lat_stab_grad_slopes_5my.png", lat.stab.grad.panels, width = 41, height = 12, scale = 0.3)
```


# Make panels with relative abundances and sliding windows
```{r}
panels = (comp_time_lat_plot /  lat.stab.grad.panels + ylab("Temporal FAVA")) + patchwork::plot_annotation(tag_levels = "A")

# ggsave("../Figures/lat_stab_grad_supplement.png", panels, width=42, height = 24, scale=0.3)
```

```{r}
# Slopes for species
species_nested_age.windows <- species_age.windows$window_data %>% 
  pivot_longer(cols = w1:w4, names_to = "index", values_to = "mya") %>% 
  transmute(pal.lat = group, FAVA, mya) %>%
  nest(data = -pal.lat) %>%
  mutate(pal.lat = pal.lat %>% stringr::str_split_i(pattern = c("[\\[,\\]]+"), 2) %>% as.numeric)

species_slopes_by_lat <- species_nested_age.windows %>%
  mutate(model = map(data, ~lm(FAVA ~ mya, data = .)))  %>%
  mutate(results = map(model, tidy)) %>%
  unnest(results)  %>% filter(term == "mya") %>% 
  transmute(pal.lat, slope=-estimate, p.value, data = "species")
# ^ I use the negative of the slope here because time is going backwards


# Slopes for ecogroups
ecogroup_nested_age.windows <- ecogroup_age.windows$window_data %>% 
  pivot_longer(cols = w1:w4, names_to = "index", values_to = "mya") %>% 
  transmute(pal.lat = group, FAVA, mya) %>%
  nest(data = -pal.lat) %>%
  mutate(pal.lat = pal.lat %>% stringr::str_split_i(pattern = c("[\\[,\\]]+"), 2) %>% as.numeric)

ecogroup_slopes_by_lat <- ecogroup_nested_age.windows %>%
  mutate(model = map(data, ~lm(FAVA ~ mya, data = .)))  %>%
  mutate(results = map(model, tidy)) %>%
  unnest(results)  %>% filter(term == "mya") %>% 
  transmute(pal.lat, slope=-estimate, p.value, data = "ecogroup")
# ^ I use the negative of the slope here because time is going backwards


# Slopes for morphogroups
morphogroup_nested_age.windows <- morphogroup_age.windows$window_data %>% 
  pivot_longer(cols = w1:w4, names_to = "index", values_to = "mya") %>% 
  transmute(pal.lat = group, FAVA, mya) %>%
  nest(data = -pal.lat) %>%
  mutate(pal.lat = pal.lat %>% stringr::str_split_i(pattern = c("[\\[,\\]]+"), 2) %>% as.numeric)

morphogroup_slopes_by_lat <- morphogroup_nested_age.windows %>%
  mutate(model = map(data, ~lm(FAVA ~ mya, data = .)))  %>%
  mutate(results = map(model, tidy)) %>%
  unnest(results)  %>% filter(term == "mya") %>% 
  transmute(pal.lat, slope=-estimate, p.value, data = "morphogroup")
# ^ I use the negative of the slope here because time is going backwards

all_slopes_by_lat = rbind(ecogroup_slopes_by_lat, 
                          morphogroup_slopes_by_lat, 
                          species_slopes_by_lat)

# Fit quadratic and abs models to each data set 
quadratic_model_species <- lm(slope ~ pal.lat + I(pal.lat^2), data = species_slopes_by_lat)
quadratic_model_ecogroup <- lm(slope ~ pal.lat + I(pal.lat^2), data = ecogroup_slopes_by_lat)
quadratic_model_morphogroup <- lm(slope ~ pal.lat + I(pal.lat^2), data = morphogroup_slopes_by_lat)

summary(quadratic_model_species)
summary(quadratic_model_ecogroup)
summary(quadratic_model_morphogroup)

stargazer::stargazer(quadratic_model_ecogroup, quadratic_model_morphogroup, quadratic_model_species)

num_fun = function(numb) formatC(numb, format = "e", digits = 2)

modelsummary::modelsummary(list(Ecogroup = quadratic_model_ecogroup, Morphogroup = quadratic_model_morphogroup, Species = quadratic_model_species), fmt = num_fun, output = "../Data_processed/quadratic_model_results.xlsx") 

# abs_model_species <- lm(slope ~ pal.lat + abs(pal.lat), data = species_slopes_by_lat)
# abs_model_ecogroup <- lm(slope ~ pal.lat + abs(pal.lat), data = ecogroup_slopes_by_lat)
# abs_model_morphogroup <- lm(slope ~ pal.lat + abs(pal.lat), data = morphogroup_slopes_by_lat)
# 
# summary(abs_model_species)
# summary(abs_model_ecogroup)
# summary(abs_model_morphogroup)
```

```{r}

slopes_plot = ggplot(all_slopes_by_lat %>% mutate(data = stringr::str_to_title(data)), aes(x = pal.lat, y = slope, color = data)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) + 
  stat_smooth(method = "glm", formula = y ~ x + I(x^2), alpha=0.3)+ 
  geom_point(size=3, alpha=0.75) +
  # geom_point(size=2, aes(color = data)) + 
  xlab("Paleo latitude") + 
  ylab("Slope:\n[change in FAVA] / [change in time (mya)]") +
  theme_bw()+ 
  scale_x_continuous(breaks = seq(-60,60,10)) + 
  scale_color_manual(values = data_pal %>% `names<-`(stringr::str_to_title(names(data_pal)))) + 
  theme(legend.title = element_blank(), legend.position = c(0.4, 0.8))

slopes_plot
# ggsave("../Figures/lat_stab_grad.png", slopes_plot, width=20, height=12, scale=0.3)


```


```{r, fig.width=10, fig.height=8}
# Calculate the predicted slopes for each latitude bin
lat.range = seq(min(all_slopes_by_lat$pal.lat), max(all_slopes_by_lat$pal.lat))

predicted_scores_quadratic <- data.frame(
  pal.lat = lat.range, 
  species = predict(quadratic_model_species, 
                    newdata = data.frame(pal.lat = lat.range)), 
  ecogroup = predict(quadratic_model_ecogroup, 
                     newdata = data.frame(pal.lat = lat.range)), 
  morphogroup = predict(quadratic_model_morphogroup, 
                        newdata = data.frame(pal.lat = lat.range))) %>% 
  pivot_longer(-pal.lat, names_to = "data", values_to = "slope")

```

I next separate each data type, so there's a bit less going on: 

```{r, fig.height=8}
slopes_plot_pvals = ggplot(all_slopes_by_lat %>% mutate(data = stringr::str_to_title(data)), aes(x = pal.lat, y = slope)) +
  geom_hline(yintercept = 0, color = "#9E9E9E") +
  geom_vline(xintercept = 0, color = "#9E9E9E") + 
  geom_line(data =predicted_scores_quadratic %>% mutate(data = stringr::str_to_title(data)), color ="black", linewidth = 1)+
  geom_point(size=3, aes(color = ifelse(p.value<0.001, "p<0.001", ifelse(p.value<0.01, "p<0.01",  ifelse(p.value<0.05, "p<0.05", ifelse(p.value<0.1, "p<0.1", "p>0.1")))))) +
  # geom_point(size=2, aes(color = data)) + 
  xlab("Paleo Latitude") + ylab("Slope:\n[change in tFAVA] / [change in time (mya)]") +
  theme_bw()+ 
  scale_color_manual(values=rev(viridis::viridis(6))[c(1:4,6)], name = "Slope\nSignificance")+
  # scale_color_viridis_c(option = "turbo") + 
  scale_x_continuous(breaks = seq(-60,60,10)) +
ggh4x::facet_grid2(.~data, 
                     strip = ggh4x::strip_themed(text_x=lapply(data_pal, function(col) element_text(color = col, face = "bold", size = 12))))+ 
  theme(strip.background = element_blank())

slopes_plot_pvals

# ggsave("../Figures/lat_stab_grad_pvals.png", slopes_plot_pvals, width=35, height=10, scale=0.3)

```

# Is there a relationship between FAVA and the sampling effort in a window for these data? 
```{r}
sampling_slopes_lat = species_abundances_5my_5deg %>% 
  pivot_longer(cols = -c(1:3), names_to = "species", values_to = "abundance") %>% 
  group_by(lat_bin) %>% summarize(total = sum(abundance)) %>% 
  left_join(all_slopes_by_lat, by = c("lat_bin" = "pal.lat")) %>% 
  filter(!is.na(slope))

sampling_plot= ggplot(sampling_slopes_lat, aes(x = total, y = slope)) + 
  geom_point(aes(fill = lat_bin), shape = 21, size = 3, alpha = 0.8) + 
  theme_bw() + 
  scale_x_log10() + 
  # stat_smooth(method="lm", alpha = 0.1, color = "grey") + 
  ggpubr::stat_cor() +
  facet_grid(~data) + 
  # scale_color_manual(values = data_pal, guide = "none") + 
  xlab("Total foraminifera abundance in paleolatitude bin") + 
    ylab("Slope:\n[change in tFAVA] / [change in time (mya)]")+ 
  scale_fill_gradientn(colors = RColorBrewer::brewer.pal(11, "Spectral"), 
                       limits = c(-66,66), 
                       name = "Paleo\nlatitude") 
sampling_plot

# ggsave("../Figures/sampling_effort_slopes.png", sampling_plot, width = 20, height = 10, scale = 0.32)

summary(lm(slope ~ log(total), data = sampling_slopes_lat %>% filter(data == "ecogroup")))
summary(lm(slope ~ log(total), data = sampling_slopes_lat %>% filter(data == "morphogroup")))
summary(lm(slope ~ log(total), data = sampling_slopes_lat %>% filter(data == "species")))
```

